---
title: 'Proposal: an online tool for cycle policy planning and public engagement'
author: "Dr. Robin Lovelace (University of Leeds), Dr. Nikolai Berkoff (independent web developer), Martin Lucas-Smith (CycleStreets Ltd)"
date: "February 7, 2015"
output:
  pdf_document:
    keep_tex: yes
  word_document: default
---

```{r, echo=FALSE, results='hide'}
pkgs <- c("png", "grid")
lapply(pkgs, library, character.only = TRUE)
```


# Introduction

The proposal is for the building of a tool which will help 
decision-makers, stakeholders and others decide the best locations to 
develop new cycle infrastructure.

The concept of this planning tool was developed during the HackMyRoute 
hackathon organised by the Leeds node of the Open Data Initiative (ODI), 
who were enthused by the possibilities it could open up.

Our objective was simple: to create the proof-of-concept of an online 
and mobile application that will provide an evidence base for policy 
makers, bicycle campaigners and the public to help better understand the 
rate of cycling in their local area.  The tool will also help 
decision-makers implement more cost-effective pro-cycling interventions 
harnessing the power of online, interactive map visualisation 
technology.


```{r, echo=FALSE}
# Todo: add link to the wicid website here
```

# Aims

The tool will allow users to visualise different metrics related to 
cycling or active travel on the map, in order to help build a strong 
case for infrastructure improvement.

For example, one proposed path may run through an area with a current 
low cycling rate that is surrounded by areas of a high cycle rate; by 
constrast, another proposed path may be in an area with an already 
medium cycle rate but have larger propensity to cycle rate with the 
connecting location. Having data that is relevant to these 
infrastructure proposals but available in a simple online tool could 
help decide which paths could deliver the most impact.  This would allow 
tapping into existing open data to create an evidence base to support 
better decision-making.

This would not simply be using currently available data as there may be 
many other metrics to which users might like to have access.  The tool 
would also have the ability to create their own custom metric based on 
the available data. For example, one metric could combine data on the 
current fastest and safest route along with the difference the new 
infrastructure would make.  We propose to create this map for Leeds.

This project will design, build, implement this online map-based tool. 
The map display will have a simple user interface allowing the user to 
select datasets of interest using heatmaps and markers which will allow 
planers have see an overview of many different factors quickly and 
simple. We propose an agile development methodology where the datasets 
of interest and layout would be subject to a process of continual 
feedback and refinement so that the end result would not be a static 
piece of software but a collaborative process between the developers and 
the planners.

The project will build on and intermesh with existing research activity 
being undertaken at the University of Leeds, and with data layer-based 
work by CycleStreets Ltd (who run the UK cycle journey planner at 
cyclestreets.net).


# Why is this different to current software?

The 'open data' revolution has greatly increased access to data. This 
can benefit transport planners, if they have access to appropriate and 
user-friendly tools, built on a strong evidence base and accessible to 
public scrutiny.

Propensity to cycle, for example, is a critical metric in deciding where 
to target new interventions and can be calculated by harnessing the flow 
data and census databases mentioned previously.

Extracting useful *information* and *knowledge* from diverse datasets 
requires specialist skills to analyse. This is a major barrier to entry 
for transport planners, policy makers and stakeholders from civil 
society. In addition, new datasets continue to emerging that could aid 
transport planner and there is a cost associated with adding this to 
current tools.


# Prototyping work

A very basic proof-of-concept prototype was developed during the 
HackMyRoute hackathon refered to above. It gives an indication of the 
levels of cycling of different parts of Leeds (as determined from the 
census) and correlates this with collision data, route accessibility, 
and cycle parking presence. The prototype can be seen at:

https://robinlovelace.shinyapps.io/fixMyPath/

The partners for the proposal include CycleStreets Ltd, a not-for-profit 
social enterprise who run a cycle journey planner and a range of tools 
for advocacy groups and local authorities. Amongst these is a 
comprehensive collision data viewer, for instance. Funding will enable 
collaboration between the partners in the project so that the combined 
strengths of the (a) research-based activity which will underpin the 
tool, and (b) the end-user reach and user knowledge available to 
CycleStreets will mean a user-friendly and widely-accessible tool.

```{r, echo=FALSE}
img <- readPNG("../../mobile-layer-example.png")
grid.raster(img)
```


# Future development proposal

\clearpage


# Appendix: data sources

The data used for this project is publicly available. We selected 
datasets based on their ability to help explain the high or low rate of 
cycling in different areas, and to guide future active travel policies, 
such as the construction of new bicycle paths and dedicated bicycle 
parking facilities. Specifically, we used the following datasets:

- Data on travel to work flows between medium super output areas
(MSOAs) --- 'origin-destination flow data'. This data has recently (summer 2014)
been made available to the public through the Open Government Licence (OGL):  .
There are 107 MSOA zones that comprise the Local Authority of Leeds.
The input dataset could easily be expanded to incorporate all 300 MSOA zones
of the West Yorkshire Combined Authority (WYCA), including
Leeds, Calderdale, Bradford, Kirklees and Wakefield.
For Leeds, on which our prototype was based, the flow data contains 10,536
origin-destination pairs. An example of this data is present in Table 1.

```{r, echo=FALSE}
# library(knitr)
# kable(fl_present[1:5, ])

```

|Origin    |Destination | Total| Bicycle| lon_origin| lat_origin|  lon_dest|
|:---------|:-----------|-----:|-------:|----------:|----------:|---------:|
|E02002330 |E02002330   |    66|       1|  -1.400577|   53.92915| -1.400577|
|E02002330 |E02002331   |   742|      38|  -1.400577|   53.92915| -1.347653|
|E02002330 |E02002332   |     2|       0|  -1.400577|   53.92915| -1.705267|
|E02002330 |E02002333   |     9|       2|  -1.400577|   53.92915| -1.687695|
|E02002330 |E02002334   |    50|       1|  -1.400577|   53.92915| -1.357652|
