---
title: '''FixMyPath'': a cycle Policy and public engagement tools'
author: "Robin Lovelace, Nikolai Berkoff, Martin Lucas-Smith"
date: "February 7, 2015"
output: pdf_document
---

# Introduction

The concept of FixMyPath was developed during the HackMyRoute hackathon organised
by the Leeds node of the Open Data Initiative (ODI). The objective was simple:
to create a proof-of-concept of an online tool application
that will provide an evidence base for policy makers,
bicycle campaigners and the public to help better understand the rate of cycling
in their local authority, and implement more cost-effective pro-cycling interventions.

# Data

The data used for this project is publicly available. We selected datasets based
on their ability to help explain the high or low rate of cycling in different
areas and to guide future pro-cycling policies, especially the construction of
new bicycle paths. Specifically, we used the following datasets:

- Data on travel to work flows between medium super output areas
(MSOAs) --- 'origin-destination flow data'. This data has recently (summer 2014)
been made available to the public through the Open Government Licence (OGL):  .
There are 107 MSOA zones that comprise the Local Authority of Leeds.
The input dataset could easily be expanded to incorporate all 300 MSOA zones
of the West Yorkshire Combined Authority (WYCA), including
Leeds, Calderdale, Bradford, Kirklees and Wakefield.
For Leeds, on which our prototype was based, the flow data contains 10,536
origin-destination pairs. An example of this data is present in Table 1.

```{r, echo=FALSE}
# library(knitr)
# kable(fl_present[1:5, ])

```

|Origin    |Destination | Total| Bicycle| lon_origin| lat_origin|  lon_dest|
|:---------|:-----------|-----:|-------:|----------:|----------:|---------:|
|E02002330 |E02002330   |    66|       1|  -1.400577|   53.92915| -1.400577|
|E02002330 |E02002331   |   742|      38|  -1.400577|   53.92915| -1.347653|
|E02002330 |E02002332   |     2|       0|  -1.400577|   53.92915| -1.705267|
|E02002330 |E02002333   |     9|       2|  -1.400577|   53.92915| -1.687695|
|E02002330 |E02002334   |    50|       1|  -1.400577|   53.92915| -1.357652|


```{r, echo=FALSE}
# Todo: add link to the wicid website here
```

- 


# Technology

# Example uses

# Future development proposal