---
title: 'Proposal: an online tool for cycle policy planning and public engagement'
author: "Robin Lovelace (University of Leeds), Nikolai Berkoff (independent web developer), Martin Lucas-Smith (CycleStreets.net)"
date: "February 7, 2015"
output:
  pdf_document:
    keep_tex: yes
  word_document: default
---

```{r, echo=FALSE, results='hide'}
pkgs <- c("png", "grid")
lapply(pkgs, library, character.only = TRUE)
```


# Introduction

The concept of this planning tool was developed during the HackMyRoute hackathon organised
by the Leeds node of the Open Data Initiative (ODI). Our objective was simple:
to create the proof-of-concept of an online application
that will provide an evidence base for policy makers,
bicycle campaigners and the public to help better understand the rate of cycling
in their local area.  The tool will also help decision makers
implement more cost-effective pro-cycling interventions harnessing the
power of online, interactive map visualisation technology.


```{r, echo=FALSE}
# Todo: add link to the wicid website here
```

# Problem

There is a clear need for a solid evidence base for prioritising new cycling
investment. (Policy stuff e.g. CDP.) At a time of fiscal restraint and increased
increased investment in active travel, 

The 'open data' revolution has greatly increased access to data.
This can benefit transport planners, if they have access to appropriate and
user-friendly tools, built on a strong evidence base and accessible to public
scrutiny.
Propensity to cycle, for example, is a critical metric in deciding where
to target new interventions and can be calculated by harnessing the
flow data and census databases mentione previously. 

In addition, new datasets continue to emerging that could aid transport planner
and there is a cost associated with adding this to current tools.

Extracting useful *information* and *knowledge* from diverse datasets 
requires specialist skills to analyse. This is a major barrier to entry for 
transport planners, policy makers and stakeholders from civil society.

# Aims

This project will design, build, implement an online map-based tool.  We will 
combine various datasets and will create a simple process for adding custom 
data to the map.  The map display will have a simple user interface
allowing the user to select datasets of interest using heatmaps and markers.
We propose an agile development methodology where the datasets of interest and layout
would be subject to a process of continual feedback and refinement.

# Example uses

https://robinlovelace.shinyapps.io/fixMyPath/

```{r, echo=FALSE}
img <- readPNG("../../mobile-layer-example.png")
grid.raster(img)
```


# Future development proposal

\clearpage

# Appendix: data sources 

The data used for this project is publicly available. We selected datasets based
on their ability to help explain the high or low rate of cycling in different
areas, and to guide future active travel policies, such as the construction of
new bicycle paths and dedicated bicycle parking facilities.
Specifically, we used the following datasets:

- Data on travel to work flows between medium super output areas
(MSOAs) --- 'origin-destination flow data'. This data has recently (summer 2014)
been made available to the public through the Open Government Licence (OGL):  .
There are 107 MSOA zones that comprise the Local Authority of Leeds.
The input dataset could easily be expanded to incorporate all 300 MSOA zones
of the West Yorkshire Combined Authority (WYCA), including
Leeds, Calderdale, Bradford, Kirklees and Wakefield.
For Leeds, on which our prototype was based, the flow data contains 10,536
origin-destination pairs. An example of this data is present in Table 1.

```{r, echo=FALSE}
# library(knitr)
# kable(fl_present[1:5, ])

```

|Origin    |Destination | Total| Bicycle| lon_origin| lat_origin|  lon_dest|
|:---------|:-----------|-----:|-------:|----------:|----------:|---------:|
|E02002330 |E02002330   |    66|       1|  -1.400577|   53.92915| -1.400577|
|E02002330 |E02002331   |   742|      38|  -1.400577|   53.92915| -1.347653|
|E02002330 |E02002332   |     2|       0|  -1.400577|   53.92915| -1.705267|
|E02002330 |E02002333   |     9|       2|  -1.400577|   53.92915| -1.687695|
|E02002330 |E02002334   |    50|       1|  -1.400577|   53.92915| -1.357652|