---
title: 'Proposal: an online tool for cycle policy planning and public engagement'
author: "Dr. Robin Lovelace (University of Leeds), Dr. Nikolai Berkoff (independent web developer), Martin Lucas-Smith (CycleStreets Ltd)"
date: "16 February, 2015"
output:
  pdf_document:
    keep_tex: yes
  word_document: default
---

```{r, echo=FALSE, results='hide'}
pkgs <- c("png", "grid")
lapply(pkgs, library, character.only = TRUE)
```


# Introduction

The proposal is to create a tool which will empower 
decision-makers, stakeholders and others to understand cycling at the local level
and to focus on the areas and routes most in need of improved cycle infrastructure.

The concept of this planning tool was developed during the HackMyRoute 
hackathon organised by the Leeds node of the Open Data Initiative (ODI).
Representatives from the ODI, West Yorkshire Combined Authority (WYCA)
and members of the public were enthused by the possibilities it could open up.
The proposed work will extend the functionality and useability of the tool,
for the benefit of all cyclists in West Yorkshire.

Our aim simple: to create an online 
and mobile application that will provide an evidence-base for understanding
variability in the rate of cycling at the local level across West Yorkshire.
The tool will provide up-to-date information on a range of factors affecting cycling
and sustainable transport policy to policy makers, bicycle campaigners and 
interested members of the public in an accessible, attractive and interactive form.

```{r, echo=FALSE}
# Todo: add link to the wicid website here
# Removed:
#   The tool could also help decision-makers implement more cost-effective pro-cycling
# interventions 
# harnessing the power of online, interactive map visualisation 
# technology.
```

# Objectives

- To create an interactive map that covers the entirety of West Yorkshire,
with information on the current rate of bicycle commuting in different areas.
- To enable users to visualise the rate of cycle commuting, not only in specific areas,
but also along 'flow lines'.
- To create metrics of the potential level of cycling in different areas, based
on the number of cyclable trips between areas on the map.
- To alocate flow lines of particular interest to the user (e.g. where the current
highest rate of cycling is or where the potential increase in cycling is highest)
onto the transport network using the CycleStreets.net API.
- To allow users to visualise addition information relevant to cycle policy
in the local area including bicycle parking, road traffic casualties and
bicycle shops. This will help priorities what type of cycling provision is
most needed in different areas.

# Example usage

The tool could be used by policy makers to identify new paths that serve areas with currently 
low cycling rates but which have a high potential for increased cycling.
Many people currently have access to the underlying data, but not presented in
a form that is easy to understand.  Thus policy makers could use the tool to 
tap into existing open data to create an evidence base to support 
better decision-making.

```{r}
# another proposed path may be in an area with an already 
# medium cycle rate but have larger propensity to cycle rate with the 
# connecting location.
## By decision makers
## By cycle campaigners

```

Another beneficiary will be
cycle campaigns. Local campaign groups
can easily become focussed on single issues, to the detriment of
understanding the 'big picture' and rates of cycling and provision in
areas outside their members' local areas. The tool could help campaigners
prioritise where to focus their next campaign, empowered with the evidence.

The public and everyday cyclist will benefit by being provided with
a simple, cohesive system that makes available a wide range of evidence on
cycling in a single place. This will inform the ongoing cycling debate in
West Yorkshire, directing it to discussion of the data rather than anecdote.

```{r}
# This would not simply be using currently available data as there may be 
# many other metrics to which users might like to have access.  The tool 
# would also have the ability to create their own custom metric based on 
# the available data. For example, one metric could combine data on the 
# current fastest and safest route along with the difference the new 
# infrastructure would make.  We propose to create this map for Leeds.
```


This project will design, build, implement this online map-based tool. 
The map display will have a simple user interface allowing the user to 
select datasets of interest using heatmaps and markers which will allow 
planers have see an overview of many different factors quickly and 
simple. We propose an agile development methodology where the datasets 
of interest and layout would be subject to a process of continual 
feedback and refinement so that the end result would not be a static 
piece of software but a collaborative process between the developers and 
the planners.

The project will build on and intermesh with existing research activity 
being undertaken at the University of Leeds, and with data layer-based 
work by CycleStreets Ltd (who run the UK cycle journey planner at 
cyclestreets.net).


# Why is this different to current software?

The 'open data' revolution has greatly increased access to data. This 
can benefit transport planners, if they have access to appropriate and 
user-friendly tools, built on a strong evidence base and accessible to 
public scrutiny.

Propensity to cycle, for example, is a critical metric in deciding where 
to target new interventions and can be calculated by harnessing the flow 
data and census databases mentioned previously.

Extracting useful *information* and *knowledge* from diverse datasets 
requires specialist skills to analyse. This is a major barrier to entry 
for transport planners, policy makers and stakeholders from civil 
society. In addition, new datasets continue to emerging that could aid 
transport planner and there is a cost associated with adding this to 
current tools.


# Prototyping work

A very basic proof-of-concept prototype was developed during the 
HackMyRoute hackathon refered to above. It gives an indication of the 
levels of cycling of different parts of Leeds (as determined from the 
census) and correlates this with collision data, route accessibility, 
and cycle parking presence. The prototype can be seen at:

https://robinlovelace.shinyapps.io/fixMyPath/

The partners for the proposal include CycleStreets Ltd, a not-for-profit 
social enterprise who run a cycle journey planner and a range of tools 
for advocacy groups and local authorities. Amongst these is a 
comprehensive collision data viewer, for instance. Funding will enable 
collaboration between the partners in the project so that the combined 
strengths of the (a) research-based activity which will underpin the 
tool, and (b) the end-user reach and user knowledge available to 
CycleStreets will mean a user-friendly and widely-accessible tool.

```{r, echo=FALSE}
img <- readPNG("../../mobile-layer-example.png")
grid.raster(img)
```


# Future development proposal

\clearpage


# Appendix: data sources

The data used for this project is publicly available. We selected 
datasets based on their ability to help explain the high or low rate of 
cycling in different areas, and to guide future active travel policies, 
such as the construction of new bicycle paths and dedicated bicycle 
parking facilities. Specifically, we used the following datasets:

- Data on travel to work flows between medium super output areas
(MSOAs) --- 'origin-destination flow data'. This data has recently (summer 2014)
been made available to the public through the Open Government Licence (OGL):  .
There are 107 MSOA zones that comprise the Local Authority of Leeds.
The input dataset could easily be expanded to incorporate all 300 MSOA zones
of the West Yorkshire Combined Authority (WYCA), including
Leeds, Calderdale, Bradford, Kirklees and Wakefield.
For Leeds, on which our prototype was based, the flow data contains 10,536
origin-destination pairs. An example of this data is present in Table 1.

```{r, echo=FALSE}
# library(knitr)
# kable(fl_present[1:5, ])

```

|Origin    |Destination | Total| Bicycle| lon_origin| lat_origin|  lon_dest|
|:---------|:-----------|-----:|-------:|----------:|----------:|---------:|
|E02002330 |E02002330   |    66|       1|  -1.400577|   53.92915| -1.400577|
|E02002330 |E02002331   |   742|      38|  -1.400577|   53.92915| -1.347653|
|E02002330 |E02002332   |     2|       0|  -1.400577|   53.92915| -1.705267|
|E02002330 |E02002333   |     9|       2|  -1.400577|   53.92915| -1.687695|
|E02002330 |E02002334   |    50|       1|  -1.400577|   53.92915| -1.357652|
